---
title: "Untitled"
output: github_document
date: "2025-06-12"
author: "Charisa Geyer"
---
# Purpose: Practicing using ggplot2 before I attempt the datsci prac project!

I am learning  how to use the RStudio add-in: GGplotAssist

# Cleaning the environment
```{r}
rm(list = ls()) # Clean your environment:
gc() # garbage collection - It can be useful to call gc after a large object has been removed, as this may prompt R to return memory to the operating system.
library(tidyverse)
list.files('code/', full.names = T, recursive = T) %>% .[grepl('.R', .)] %>% as.list() %>% walk(~source(.))
```
# Installing the prerequisite and add-in package
```{r}
#install.packages("devtools")
devtools::install_github("cardiomoon/editData")
#install.packages("devtools")
devtools::install_github("cardiomoon/ggplotAssist")
```
# Loading the dataset to start with
```{r}
library(ggplot2)
library(ggplotAssist)
data_use <- fmxdat::Indexes
fmxdat::Indexes_Labs

joined_data_use <- left_join(fmxdat::Indexes %>% gather(Country, TRI, -Date), 
fmxdat::Indexes_Labs,
by = "Country") %>% arrange(Date)

# highlight the dataframe/tibble and click on AddIns -> GGplotAssist
```
# Now let's follow the tutorial given in class

## Tidy up the data
```{r}
pacman::p_load(tidyverse)

plot_df <- 
left_join(
fmxdat::Indexes %>% gather(Country, TRI, -Date),
fmxdat::Indexes_Labs,
by = "Country"
) %>% mutate(Label = coalesce(Label, "Other")) %>% arrange(Date)
  
print(plot_df)
```

## Plot all Brics countries together in one plot
```{r}
dfPlotData <- plot_df %>% filter(Label == "Brics")
g <- 
  dfPlotData %>% 
  
  # Initialize the canvas:
  ggplot() +
  
  # Now add the aesthetics:
  geom_line(aes(x = Date, y = TRI, color = Country), alpha = 0.8, size = 1) 

# Notice that I used alpha to adjust the line opacity (useful when having overlapping lines e.g.)

print(g)
```
```{r}
# Create smaller dataframe with only the Brics countries
dfPlotData <- plot_df %>% filter(Label == "Brics")

# Create the plot list
g <- dfPlotData %>% 
    ggplot() + # Create the canvas
    geom_line(aes(x = Date, y = TRI, color = Country), alpha = 0.8, size = 1)
print(g)

```


Now bring g down (the plot) and add a couple of things: 
```{r}
g <- 
  g + 
  theme_bw() + 
  theme( legend.position = "bottom") + 
  labs(x = "", y = "Prices",
       title = "TRI of Brics Countries", 
       subtitle = "Total Return Index using net dividends reinvested",
       caption = "Note:\nBloomberg data used")
  
g
```

## Breaking up each line into its own plot

```{r}
g <- # redefine g now!
g +
  facet_wrap(~Country)

g
```

Almost there… let’s scale the axes to be free for each y, and also change the colours a bit using ggthemes:
```{r}
pacman::p_load(ggthemes)
g <- 
g + 
  facet_wrap(~Country, scales = "free_y") + 
  ggthemes::scale_color_wsj()

g
```

Another theme library to consider is ggsci. * Here’s the gallery, and you can use this as shown here.

Let’s use e.g. npg as color, and also change the thickness of the lines (see size parameter…)
```{r}
pacman::p_load("ggsci")
g + scale_color_npg()
```
Can do a deeper analysis of a subset of dates etc, since the "figures are alive"
```{r}
g %+% subset(dfPlotData, Date > as.Date("2009-01-31")) + labs(title = "Post-Crisis TRI")
```
```{r}
g %+% subset(dfPlotData, Date < as.Date("2009-01-31")) + labs(title = "Pre-Crisis TRI")
```
You could also retrieve the data of a ggplot object simply as:
```{r}
Plot_Data_Recovered <- g$data
```

You could also add another dataframe and some labels to your plot as:


```{r}
pacman::p_load(lubridate)
Another_df <- fmxdat::findata %>% gather(Stocks, Px, -Date) %>% filter(Stocks == "JSE.SLM.Close")

g + 
  geom_line(data = Another_df, aes(Date, Px, color = Stocks), colour = "steelblue", alpha = 0.3) + 
  geom_label(data = Another_df %>% filter(Date == last(Date)), aes(Date, Px, label = Stocks), size = 3, alpha = 0.1) + 
  
  geom_label(data = dfPlotData %>% filter(Date == last(Date)), aes(Date, TRI, label = Country), size = 3, alpha = 0.1) 
```

# Saving your plots
```{r}
rmsfuns::build_path( paste0( getwd(), "Figures"))
ggsave(
  filename = "Figures/Plot.png", # here you can also give your figure a name!
  plot = g, width = 6, height = 6, device = "png")
```

# Other plots

#Density plots
```{r}
plot_data <- 
  dfPlotData %>% 
  
  group_by(Country) %>% 
  
  mutate(Return = TRI/lag(TRI)-1) %>% ungroup()

gdens <- 
  plot_data %>% 
  ggplot() + 
  geom_density(aes( x = Return, fill = Country )) #Note - here we set the fill to be given by ‘Country’. Fill works similarly to colour, where the latter would’ve simply coloured the line, not the area under the curve.

gdens
```

### Adjust opacitity

### Colour the lines

```{r}
gdens <- 
gdens +
  
  geom_density(aes( x = Return, fill = Country , colour = Country), alpha = 0.2, size = 1.25) + 
  
  ggthemes::theme_economist_white()

gdens
```

Subsetting to look only at the deep left-tail (say returns below 10%) is also super easy:

Notice that subset requires a logical operator (i.e. must give TRUE / FALSE). The following won’t work therefore as it is a wrangle, not a logical operator:
```{r}
#gdens %+% subset(plot_data, ifelse(Return < -0.1, -0.1, Return) ) + labs(title = "Left tail plot")
```

So, let’s bring the tails in a bit to not skew the graph as much. This requires a re-wrangling, not simply a subsetting:

Let’s trim our returns to be between 5% - 95% quantiles of returns only:

# Winzorising

```{r}
# Winzorising to be between 5% and 95% can be done as follows:

new_plot_data <-
    plot_data %>% 
    filter(!is.na(Return)) %>% group_by(Country) %>% 
    mutate(q05 = quantile(Return, na.rm=T, probs=0.05), q95 = quantile(Return, na.rm=T, probs = 0.95), Return = ifelse(Return >= q95, q95, ifelse(Return <= q05, q05, Return))) %>% 
    
    select(-starts_with("q")) %>% 
    ungroup()

plot_data <- 
  plot_data %>% 
  
  filter(!is.na(Return)) %>% group_by(Country) %>% 
  
  mutate(q05 = quantile(Return, na.rm=T, probs = 0.05), q95 = quantile(Return, na.rm=T, probs = 0.95), Return = ifelse(Return >= q95, q95, ifelse(Return <= q05, q05, Return))) %>% 
  
  select(-starts_with("q")) %>% ungroup()

gdens <- 
  ggplot(plot_data) +
  
  geom_density(aes( x = Return, fill = Country , colour = Country), alpha = 0.2, size = 1.25) + 
  
  ggthemes::theme_economist_white() + 
  
  labs(y = "Density (winsorized)", x = "Return", title = "Return density plot") + 
  
  theme(legend.position = "bottom")

gdens
```
# Figures in functions
Of course! Ideally you should specify templates for your functions and source it like you would any other comment.

E.g., in the following code chunk I introduce you to boxplotting and jitter plotting combined in ggplot. I do this in a function, where I:

Make my function safe to illustrate an intentional error with a clear message to solve it

Show you how a figure’s design can be flexibly incorporated into a function:

```{r}
# Create this function and sve it in code file
# pacman::p_load(purrr)
# 
# Jitter_Boxplot <- function(data_frame, Title, Subtitle, Caption, Xlab, Ylab, Alpha_Set = 0.5) {
# 
#   # Dataframe should be tidy, and be of the form: 
#       # date  | Ticker  |   Identifier  |  Return  
# 
#   if( !"Identifyer" %in% names(data_frame) ) stop("\n\nERROR:::::>Please provide valid Identifier column!\n\n")
#   if( !"Date" %in% names(data_frame) ) stop("Please provide valid Date column ")
#   if( class(data_frame$Date) != "Date" ) stop("Date column not of class Date ")
#   
# g1 <-   
# ggplot(data_frame) + 
#   
#   geom_boxplot(aes(x = Identifier, y = Return, fill = Country), alpha = Alpha_Set) + 
#   
#   geom_jitter(aes(x = Identifier, y = Return, color = Country, alpha = Alpha_Set)) + 
#   
#     theme_bw() + 
#   
#   guides(color = FALSE, fill = FALSE, alpha = FALSE) +
#   # Add titles:
#   labs(title = Title, 
#        subtitle = Subtitle,
#        caption = Caption,
#        x = Xlab, y = Ylab) + 
#   
#   scale_color_npg() + # Now we use fill...
#   
#   scale_fill_npg() # Now we use fill...
# 
# g1  
# 
# }


SafeJitter <- purrr::safely(Jitter_Boxplot)
# Let's first break it to see the error produced:
Result <-SafeJitter(data_frame = plot_data, 
               Title = "BRICS Return Histograms", 
               Subtitle = "Transparency is key", 
               Caption = "Data was downloaded from Bloomberg", 
               Xlab = "", Ylab = "Distribution",
               Alpha_Set = 0.4)

  print(Result$error)
```

```{r}
Result <- 
SafeJitter(data_frame = plot_data %>% mutate(Identifyer = Country), 
               Title = "BRICS Return Histograms", 
               Subtitle = "Transparency is key", 
               Caption = "Data was downloaded from Bloomberg", 
               Xlab = "", Ylab = "Distribution")
  
g <- Result$result

g
```

# Point plots
Let’s create a mean-variance plot of various crypto currencies over the last 52 weeks - with the size of the bubble in the plot given by its market at the last date.

(Notice this requires a bit of planning and wrangling before we plot)

```{r}
Bub_Size <- 
  fmxdat::cryptos %>% filter(date == last(date)) %>% 
  
  select(name, market) %>% rename(Size = market) %>% unique

MeanVar <- 
  fmxdat::cryptos %>% 
  
  filter(date > last(date) %m-% weeks(52)) %>% 
  
  group_by(name) %>% mutate(Return = close / lag(close) - 1) %>% 
  
  summarise_at(vars(Return), list(Mean = ~mean(., na.rm = T), SD = ~sd(., na.rm=T)))

# Let's append the size column:
plot_df <- left_join(MeanVar, Bub_Size, by = "name")

g <- 
  plot_df %>% ggplot() + 
  
  geom_point(aes(Mean, SD, size = Size), color = "steelblue", alpha = 0.6) + guides(size = F) + 
  
  theme_bw() + 
  
  labs(title = "Terrible scales")
g

```
Fixing the scale a bit

```{r}
plot_df <- 
  
  plot_df %>% 
  
  mutate(Mean = ifelse(Mean > 0.15, 0.15, Mean)) %>% 
  
  mutate(SD = ifelse(SD > 0.25, 0.25, SD))

g <- 
  plot_df %>% ggplot() + 
  
  geom_point(aes(Mean, SD, size = Size), color = "steelblue", alpha = 0.6) + guides(size = F) + 
  
  theme_bw() + 
  
  labs(title = "Terrible Size Spread")
g
```
still not so happy with that, let's set quantiles rather
Notice my setting of colours based on a column, as well as size based on a value above. Also - note how I changed the legend title
```{r}
plot_df_Size_Adj <- 
  plot_df %>% 
  mutate(Q1 = quantile(Size, 0.1)) %>% 
  
  mutate(Q2 = quantile(Size, 0.8)) %>%
  
  mutate(Size = ifelse(Size < Q1, 5,
                       ifelse(Size < Q2, 6,
                              ifelse(Size >= Q2, 7, NA_real_ )))) %>% 
  
  mutate(ColorSet = ifelse(Size == 5, "small", ifelse(Size == 6, "medium", "large")))

g <- 
  plot_df_Size_Adj %>% ggplot() + 
  
  geom_point(aes(Mean, SD, size = Size, color = ColorSet), alpha = 0.6) +
  
  scale_color_manual(values = c(small = "darkgreen", medium = "steelblue", large = "darkred")) + 
  
  guides(size = F, color=guide_legend(title="Size Guide")) + 
  
  theme_bw() + 
  
  labs(title = "Size and spread sorted")
g
```

# Barplots
 To produce barplots in ggplot you have to be somewhat careful, the notation can initially be a bit confusing.

I give you an example below of a simple bar plot:

For this I show you a few tricks:
Notice how I transform a numeric column into a date column

Notice how I calculate the max wind per year - after which I calculate the percentage of that year’s max wind relative to all time max. Same for pressure

In the plot code - notice how I make the axes show only every second year, and doing so show the month and year as e.g. Jan ’82

Notice how I flip the x-axes to be vertical.

Notice that the legends are dropped as they are redundant as the facet title are descriptive enough.
```{r}
df_barplot <-
  
  dplyr::storms %>% 
  
  group_by(year) %>% summarize(max_wind = max(wind), max_pressure = max(pressure)) %>% 
  
  ungroup() %>% 
  
  mutate(max_wind = max_wind / max(max_wind), max_pressure = max_pressure / max(max_pressure)) %>% 
  # Let's make it tidy...
  gather(Type, Value, -year) %>% 
  # Let's make column year a valid date object:
  mutate(year = as.Date( paste0( year, "-01-01" )) )


df_barplot %>% 
  
  ggplot() + 
  
  geom_bar(aes(x = year, y = Value, fill = Type), stat = "identity") + 
  # scale_
  facet_wrap(~Type, scales = "free_y", nrow = 2) + 
  
  scale_x_date(labels=scales::date_format("%b '%y"), date_breaks = "2 years") + 
  
  theme_bw() + 
  
    theme(axis.text.x=element_text(angle = 90, hjust = 1)) + 
  
  labs(x = "", y = "") + guides(fill = F)
```

# Star wars plotting
For ordering plots, we make use of the forcats package in the code below. First, let’s plot the frequency of eye-colours for people in Star Wars:
```{r}
pacman::p_load(forcats)

SW <- 
  dplyr::starwars %>% 
  
  select(name, height, mass, hair_color, gender, eye_color) %>% 
  
  mutate(height = as.numeric(height))

SW %>%
  
  mutate(eye_color = forcats::fct_infreq(eye_color)) %>%
  
  ggplot() + 
  
  geom_bar(aes(x = eye_color), fill = "steelblue", alpha = 0.7) + 
  
  coord_flip() + 
  
  labs(x = "Count", y = "Eye-Colour", title = "Eye-Colour spread in Star Wars movies", caption = "Data from dplyr package in R")
```
Let’s plot the height of male characters in Starwars, arranged by height, with a dotted line indicating the height of Darth Vader:
1. only males
2. height
3. arrange in order by height
4. add dotted line for height of DV

filter(Date > first(Date)) %>%


```{r}
height_plot <- SW %>% 
    
    filter(!is.na(height)) %>%
    filter(gender == "masculine") %>%
    mutate(name = as_factor(name)) %>% 
    mutate(name = fct_reorder(name, height)) %>% # This **converts** the `name` column from a character vector into a **factor**
    mutate(DVaderheight = max(ifelse(name == "Darth Vader", height, NA_real_), na.rm = T)) %>%
    arrange(height)  

ggplot(height_plot, aes(x = name, y = height)) +
    geom_col() + # must say this to get the cols to show up! 
    geom_hline(aes(yintercept = DVaderheight), linetype = "dotted", color = "red", size = 1) +
    # coord_flip() +
    labs(title = "Character heights in Starwars", y = "Height in cm", x = "Character name") +
    theme_economist_white() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```


```{r}
height_plot <- SW %>%
  filter(!is.na(height)) %>%
  filter(gender == "masculine") %>%
  mutate(name = as_factor(name)) %>%
  mutate(name = fct_reorder(name, height)) %>%
  mutate(DVaderheight = max(ifelse(name == "Darth Vader", height, NA_real_), na.rm = TRUE)) %>%
  arrange(height)

height_plot %>% 
    ggplot() +
    geom_bar(aes(x = name, y = height, color = "steelblue"), stat = "identity") +
   # scale_fill_manual(values = "darkgreen") +
    geom_hline(aes(yintercept = DVaderheight), linetype = "dotted", color = "red", alpha = 0.8, size = 1) +
    theme_bw()+
    geom_label( data = height_plot %>% filter(height == median(height, na.rm=T)) 
                %>% head(1), aes(name, DVaderheight, label = glue::glue("Vader's height: {max(DVaderheight)} cm")), alpha = 0.5) +
    guides(fill = F) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust =1)) +
    labs(title = "Heights of Characters in StarWars", x = "Character name", y = "Height in cm")


ggplot(height_plot, aes(x = name, y = height)) +   # No reorder() needed again
  geom_col() +                                     # Use geom_col (like geom_bar with stat = "identity")
  geom_hline(aes(yintercept = DVaderheight),       # Correct function to use
             linetype = "dotted", color = "red", size = 1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(title = "Character heights in Star Wars",
       y = "Height in cm", x = "Character name")
    
```

```{r}
plot_df <- 
  
SW %>% filter(!is.na(height)) %>% 
  
  filter(gender == "masculine") %>% 
  
  mutate(name = as_factor(name)) %>% 
  
  mutate(name = fct_reorder(name, height)) %>% 
  
  mutate(VaderHeight = max( ifelse(name == "Darth Vader", height, NA_real_), na.rm = T ) ) %>% 
  
  mutate(Taller_Than_Vader = ifelse(height > VaderHeight, "Taller", "Shorter"))
  
  plot_df %>% 
    
    ggplot() + 
    
    geom_bar(aes(name, height, fill = Taller_Than_Vader), stat= 'identity') + 
    
    scale_fill_manual(values = c(Taller = "darkred", Shorter = "darkgreen")) + 
    
    theme_bw() + 
    
    theme(axis.text.x=element_text(angle = 90, hjust = 1)) + 
    
    geom_hline(aes(yintercept = max(VaderHeight)), linetype = "dotted", size = 2, alpha = 0.8, color = "steelblue") + 
    
    geom_label( data = plot_df %>% filter(height == median(height, na.rm=T)) %>% head(1), aes(name, VaderHeight, label = glue::glue("Vader's height: {max(VaderHeight)} cm")), alpha = 0.5) + 
    guides(fill = F) + 
    labs(title = "Height of Star Wars characters relative to Darth Vader's height", caption = "Data from dplyr package",
         subtitle = "Seems Darth Vader is not the tallest....")
```

# Some bonus examples
Combining multiple plots onto one: 
```{r}
pacman::p_load(patchwork)

df_plot <- 
  
left_join(
  
fmxdat::Indexes %>% gather(Country, TRI, -Date),

fmxdat::Indexes_Labs,

by = "Country"

) %>% 
  
  mutate(Label = coalesce(Label, "Other")) %>% 
  
  arrange(Date)


p1 <- 
  df_plot %>% 
  
  filter(Label == "Brics") %>% 
  
  ggplot() + 
  
  geom_line(aes(Date, TRI, color = Country)) + 
  
  theme_bw() + 
  
  theme(legend.position = "bottom")

p2 <- 
SafeJitter(data_frame = plot_data %>% mutate(Identifyer = Country), 
           
               Title = "BRICS Return Histograms", 
           
               Subtitle = "Transparency is key", 
           
               Caption = "Data was downloaded from Bloomberg", 
           
               Xlab = "", Ylab = "Distribution")$result

p3 <- 
    plot_df %>% 
  
    ggplot() + 
  
    geom_bar(aes(name, height, fill = Taller_Than_Vader), stat= 'identity') + 
  
    scale_fill_manual(values = c(Taller = "darkred", Shorter = "darkgreen")) + 
  
    theme_bw() + 
  
    theme(axis.text.x=element_text(angle = 90, hjust = 1)) + 
  
    geom_hline(aes(yintercept = max(VaderHeight)), linetype = "dotted", size = 2, alpha = 0.8, color = "steelblue") + 
  
    geom_label( data = plot_df %>% filter(height == median(height, na.rm=T)) %>% head(1), aes(name, VaderHeight, label = glue::glue("Vader's height: {max(VaderHeight)} cm")), alpha = 0.5) + 
    guides(fill = F)

p3/ (p2 +p1) + 
  plot_annotation(
  title = 'Patching figures together',
  caption = 'Source: @littlemissdata'
) & 
  theme(text = element_text('bold'))
```
# Embed images into tidyverse
Second, how to embed a static graphic (jpg / png) into a tidyframe:
An easy way though is to use the magick & ggpubr packages as follows:


```{r}
pacman::p_load(cowplot, magick, ggpubr)
Img_embed <- image_read("https://i.imgur.com/eDCUyUql.jpg")

Img_embed <- 
  
  ggplot() +
  
  background_image(Img_embed) + coord_fixed()

Img_embed / p3 +   plot_layout(widths = c(1, 1))
```
# Examples
## US Murders
Let’s create a plot for each region (South, West, North East) that:

shows the percentage murders for every state per 1 million people;

Show the Name and % text of the best and worst states (tip: use ggrepel to not have text boxes overlap);

Ensure the four plots have the same axes

Ensure axes are log-scaled

Have size of bubbles be reflective of amount of murders.

get data from dslabs

```{r}
#pacman::p_load("dslabs", "tidyverse", "ggthemes", "ggrepel", "scales")

murdr <- 
  
  dslabs::murders %>%
  
  mutate(mrate = total / population * 10^6) %>% ## new col with rate of murder per million 10^6
  
  mutate(pop_print = population / 1e6) ## just show the pop in another way


murdr %>% 
  ggplot() +
  
  geom_point(aes(pop_print, mrate, size = total, color = region)) + 
  
  facet_wrap(~region) + ## by region ? 
  
  scale_x_log10() +
  
  scale_y_log10() +
  
  theme_bw() + 
  
  # Let's alter the strip text and background a bit:
  theme(strip.background = element_rect(fill = "steelblue"),
        strip.text = element_text(face = "bold", colour = "black", size = 10)) + 
  
  scale_color_manual(values = c("red", "blue", "darkgreen", "orange")) + 
  
  labs(title = "Murders per Million by Region", x = "Population million (Log Scaled)",
       y = "Murders per million (Log Scaled)") + 
  
  ggrepel::geom_label_repel(data = murdr %>% group_by(region) %>% arrange(mrate) %>% filter(mrate == max(mrate)), aes(pop_print, mrate, label = glue::glue("{state}:\n{round(mrate, 1)} per mn")), size = 4, alpha = 0.35, fill = "red") + ## no automatic overlapping!!
  
  ggrepel::geom_label_repel(data = murdr %>% group_by(region) %>% arrange(mrate) %>% filter(mrate == min(mrate)), aes(pop_print, mrate, label = glue::glue("{state}:\n {round(mrate, 1)} per mn")), size = 4, alpha = 0.35, fill = "green") + ## \n means newline
  
  guides(color = F) + 
  
  theme(legend.title = element_text("Total Murders"))
```















## 2016 US Election polls
## 